---
title: "Probability Distributions"
---

# Functions for basic satisical models.

As an example, let $X$ be a normally distributed  random variable, $X \sim N(\mu,\sigma^2)$.
The following functions are included in R order to work with the normal distribution:

* **<tt>dnorm(x,mean=$\mu$,sd=$\sigma$)</tt>**, which returns the value of the probability density function in  **<tt>x</tt>** for the normal distribution;
* **<tt>pnorm(q,mean=$\mu$,sd=$\sigma$,lower.tail=TRUE)</tt>**, which returns $P(X\leq q)$ when lower.tail is set to TRUE, otherwise $P(X\geq q)$;
* **<tt>qnorm(p,mean=$\mu$,sd=$\sigma$)</tt>**, which calculates the theatrical quantile for a given probability level  **<tt>p</tt>**;
* **<tt>rnorm(n,mean=$\mu$,sd=$\sigma$)</tt>**, which generates n observation from a normal distribution.  

These concepts generally hold true for all the distribution functions built into R. Indeed, we will use **<tt>rbinom()</tt>** to generate observation from a binomial distribution. Use `help(distributions)` to discover all the other functions the others distributions.

# Exercises

## Exercise 1
Look at the three histograms.
```{r ex1,fig.align='center',out.width = '80%'}
par(mfrow=c(2,2))
## Random sample of size n=10 of N(0,1)-Distribution
x1 = rnorm(n = 10, mean = 0, sd = 1)	
hist(x1, breaks=3)
## Random sample of size n=100 of N(0,1)-Distribution
x2 = rnorm(n = 100, mean = 0, sd = 1)	
hist(x2, breaks=3)
## Random sample of size n=1000 of N(0,1)-Distribution
x3 = rnorm(n = 1000, mean = 0, sd = 1)	
hist(x3, breaks = 3)
```

What is a reasonable choice of the parameter **<tt>breaks</tt>**? Obtain the following graph for the random sample **<tt>x3</tt>**.
<div class='fold s'>
```{r fig.align='center',out.width = '80%'}
x3 = rnorm(n = 1000, mean = 0, sd = 1)
hist(x3, breaks = 50,probability=TRUE,xlab =NA, main='Standard normal distribution')
lines(sort(x3),dnorm(sort(x3)),col='red')
```
</div>

## Exercise 2
Assume, the survival probability of patients after brain surgery is exponentially distributed. Assume further that the average survival time of a patient is 36 months, $\lambda = 1/36$.

Plot the density function of the exponential distribution over the interval  **<tt>[0,200]</tt>**. Plot the function over a sequence obtained using the **<tt>seq()</tt>** function.

<div class='fold s o'>
```{r fig.align='center',out.width = '80%'}
x = seq(0, 200, length.out = 100)
y = dexp(x, rate = 1/36)
plot(x, y, type = "l", xlab = "months")
```
</div>

Plot the cumulative distribution function of the exponential distribution.
<div class='fold s o'>
```{r fig.align='center',out.width = '80%'}
x = seq(0, 200, length.out = 100)
y = pexp(x, rate = 1/36)
plot(x, y, type = "l", xlab = "months")
```
</div>

Determine the probability of a particular patient of surviving equal or shorter than 12 month, namely $P(X\leq 12)$.

<div class='fold s o'>
```{r}
pexp(q=12, rate=1/36)
```
</div>

What is the probability of a particular patient of surviving longer than 5 years (i.e. 60 month)?

<div class='fold s o'>
```{r}
1 - pexp(q=60, rate=1/36) # equal to pexp(q=60, rate=1/36, lower.tail = FALSE) 
```
</div>

The previous questions are equivalent to calculate the red and blue area in following plot.

```{r echo=FALSE,fig.align='center',out.width = '80%'}
curve(dexp(x,rate=1/36),from=0,to=200,xlab="months",ylab='Density')
cord.x<-c(60,seq(60,200,0.1),200)
cord.y<-c(0,dexp(seq(60,200,0.1),rate=1/36),0)
polygon(cord.x,cord.y,col='skyblue')

cord.x<-c(0,seq(0,12,0.1),12)
cord.y<-c(0,dexp(seq(0,12,0.1),rate=1/36),0)
polygon(cord.x,cord.y,col='coral1')
```



## Exercise 3
Let's assume that the outcome of a therapy can be modeled as a binomial distribution , $B(n,p)$. If $n=10$ and $p=0.7$, determine the probability that $k = \{1, 4, 8\}$ OR LESS patients out of $n = 10$ will have a successful therapy.

<div class='fold s o'>
```{r results='hold'}
pbinom(1, 10, prob=0.7)
pbinom(q=4, size=10, prob=0.7)
pbinom(q=8, size=10, prob=0.7)
pbinom(q=c(1, 4, 8), size=10, prob=0.7)
```
</div>

Determine the probability that EXACTLY $k = \{1, 4, 8\}$ patients out of $n = 10$ will have a successful therapy.

<div class='fold s o'>
```{r results='hold'}
dbinom(x=1, size=10, prob=0.7)
dbinom(x=4, size=10, prob=0.7)
dbinom(x=8, size=10, prob=0.7)
dbinom(x=c(1, 4, 8), size=10, prob=0.7)
```
</div>
