---
title: "Hypothesis Tests"
---

```{r, code=readLines('_setup.R')}
```

# Exercises

## Exercise 4.1
Consider CBF-levels were measured in 100 subjects with low-grade tumors and 100 subjects with high-grade tumors. Here we simulate the measurements from the normal distribution.

```{r ex3.1, echo=TRUE}
set.seed(12) # we set the seed to get a reproducible random result.
n = 100 
low = rnorm(n = n, mean = 50, sd = 10)
high = rnorm(n = n, mean = 60, sd = 10)
```

In a real-world experiment you would first check whether your data is normally distributed (necessary for t-test). Draw quantile-quantile plots  (QQ-plots) to check the normality-assumption (**<tt>qqnorrm</tt>**,**<tt>qqline</tt>**).

<div class='fold s o'>
```{r,eval=op_eval,echo=op_echo}
par(mfrow=c(1,2))
qqnorm(low,main='low')
qqline(low)
qqnorm(high,main='high')
qqline(high)
```
</div>

Illustrate the sample values by box-plots and histograms to check whether the assumption of equal variances seems reasonable.

<div class='fold s o'>
```{r,eval=op_eval,echo=op_echo}
boxplot(low, high, names=c("low", "high"),cex.lab=1.5, cex.axis=1.5,col=c(2,3))
grid()
hist(low, border=2, xlim=c(0, 100), ylim=c(0, 0.5*n), density=5, col=2)
hist(high, border=3, add=TRUE, density=2.5, col=3)
```
</div>

Perform the t-test (**<tt>t.test</tt>**).

<div class='fold s o'>
```{r,eval=op_eval,echo=op_echo}
a = t.test(low, high)
a = t.test(low, high, var.equal = TRUE)
#a = t.test(low, high, paired = FALSE,
#                      alternative = "two-sided",
#                      conf.level = 0.95)
a
# mean difference
a$estimate[1] - a$estimate[2]
```
</div>

Interpret the results.


## Exercise 4.2

Repeat the above computer simulation with sample sizes $n=5$ per group. Analyze whether  the test results lead to a false positive or false negative decision.

<div class='fold s o'>
```{r,eval=op_eval,echo=op_echo}
set.seed(1)
n = 5 ### Sample Size
low = rnorm(n=n, mean=50, sd=10)
high = rnorm(n=n, mean=60, sd=10)
t.test(low, high, var.equal = TRUE)
```
</div>



## Exercise 4.3
Load the data frame in the **<tt>dimarta_trial.csv</tt>** file.
```{r}
dimarta_df<-read.csv('./data/dimarta_trial.csv',sep=',')
str(dimarta_df)
```

Compute the difference between **<tt>histamine_end</tt>** level and **<tt>histamine_start</tt>** level and add the data to the data frame.

<div class='fold s'>
```{r,eval=op_eval,echo=op_echo}
dimarta_df$histamine_change<-with(dimarta_df,histamine_end - histamine_start)
```
</div>

Use boxplots to check whether the assumption of equal variances of histamine changes between arms seems reasonable .

<div class='fold s o'>
```{r,eval=op_eval,echo=op_echo}
armI<-subset(dimarta_df,arm=='I')$histamine_change
armII<-subset(dimarta_df,arm=='II')$histamine_change
armIII<-subset(dimarta_df,arm=='III')$histamine_change
boxplot(armI,armII,armIII,names=c('arm I','arm II','arm III'),
        col=c('coral1','lightskyblue','mediumspringgreen'))
grid()
```
</div>

Check whether your data is normally distributed (necessary for t-test). Draw quantile-quantile plots (QQ-plots) to check the normality-assumption,(**<tt>qqnorrm</tt>**,**<tt>qqline</tt>**).

<div class='fold s o'>
```{r,eval=op_eval,echo=op_echo}
par(mfrow=c(1,3))
qqnorm(armI,main='arm I',col='coral1')
qqline(armI)
qqnorm(armII,main='arm II',col='lightskyblue',ylab=NA)
qqline(armII)
qqnorm(armIII,main='arm III',col='mediumspringgreen',ylab=NA)
qqline(armIII)
```
</div>

To check this you can also use the Shapiro-Wilk’s test, where The null hypothesis is that the data are normally distributed (**<tt>shapiro.test</tt>**).

<div class='fold s o'>
```{r collapse=TRUE,eval=op_eval,echo=op_echo}
shapiro.test(armI)
shapiro.test(armII)
shapiro.test(armIII)
```
</div>

Let treatment arm I be the control arm and II and III two new types of drugs. Check if the population means of arm I and II are equal, then  do the same for arm I and III (**<tt>t.test</tt>**).

<div class='fold s o'>
```{r,collapse=TRUE,eval=op_eval,echo=op_echo}
#when the variance are not equal we the Welch t-test
t.test(armI,armII)
t.test(armI,armII,var.equal = TRUE)
t.test(armI,armIII)
t.test(armI,armIII,var.equal = TRUE)
```
</div>

Interpret the results of the test. 










