---
title: "Hypothesis Tests: Non-normal"
---

# Install an R package
R packages are a collection of R functions, complied code and sample data. We install an R package through 
the function **<tt>install.package(package_name)</tt>**. Let's install the package **<tt>medicaldata</tt>**:


```{r ,eval=FALSE}
install.packages("medicaldata")
```

To use this package in our session we need to load it. To accomplish this, we use the **<tt>library()</tt>** function:

```{r}
library(medicaldata)
```

from this package we want to retrieve the **<tt>polyps</tt>** data frame, which contains the systolic blood pressure, the weight and the smoking status for a sample of 500 adults. We use the **<tt>library()</tt>** to add this data frame to variable in our current environment:

```{r, collapse=TRUE}
data(polyps)
str(polyps)
```

The database **<tt>polyps</tt>** contains the results of the randomized controlled trial of sulindac vs. placebo in the prevention of colonic polyps in FAP (Familial Adenomatous Polyposis). You can find more information here [polyps](https://htmlpreview.github.io/?https://github.com/higgi13425/medicaldata/blob/master/man/description_docs/polyps_desc.html).



# Exercises

## Exercise 1 
Assume the expression of a particular gene is measures in blood samples of $n=10$ patients with Parkinson's disease and in $n=10$ healthy controls.

Example data is drawn from exponential distributions as follows:

```{r ex5.1, echo=TRUE}
set.seed(1223)
n = 10
patients = rexp(n = n, rate = 1/2)
controls = rexp(n = n, rate = 1/1)
```

Plot the sample values:
<div class='fold s o'>
```{r fig.align='center',out.width = '70%'}
complete = c(patients, controls)
group = c(rep(1, n), rep(2, n))
plot(group, complete, cex=1.5, xlim=c(0.5, 2.5), axes=FALSE, ylab="gene expression")
axis(1, c(1, 2), c("patients", "control"))
axis(2)
box()
```
</div>

Calculate the ranks of sample values. Like above, plot the ranks of sample values.
<div class='fold s o'>
```{r,fig.align='center',out.width = '70%'}
ranked.values = rank(complete)
group = c(rep(1, n), rep(2, n))
plot(group, ranked.values, cex=1.5, xlim=c(0.5, 2.5), axes=FALSE, ylab="gene expression")
axis(1, c(1, 2), c("patients", "control"))
axis(2)
box()
```
</div>

Use quantile-quantile plots to check for the assumption of normal distribution.

<div class='fold s o'>
```{r,fig.align='center',out.width = '70%'}
par(mfrow=c(1,2))
qqnorm(patients,main = 'patients')
qqline(patients)
qqnorm(controls,main='control',ylab=NA)
qqline(controls)
```
</div>

Perform Mann-Whitney U test (Wilcoxon Rank Sum test, R function (**<tt>wilcox.test</tt>**). Conduct a t-test. Compare results from these two tests.

<div class='fold s o'>
```{r, collapse=TRUE}

#Null hypothesis is that the distributions of x and y differ by a
#location shift of mu
wilcox.test(patients,controls)

#Comparison with t-test result, null hypothesis  equal means
t.test(patients, controls)
```
</div>

<!-- we can not trust t-test--> 





## Exercise 2

Compute the difference between **<tt>number12m</tt>** level and **<tt>baseline</tt>** level and add the data to the data frame.

<div class='fold s o'>
```{r}
polyps$change12m<-polyps$number12m-polyps$baseline
```
</div>

Check if the differences relative to the `treatment=='placebo'` are normally distributed and do the same for `treatment=='sulindac'`.

<div class='fold s o'>
```{r ,fig.align='center',out.width = '70%',collapse=TRUE}
dp<-subset(polyps,treatment=='placebo')$change12m
ds<-subset(polyps,treatment=='sulindac')$change12m

#Shapiro-Wilk’s test
shapiro.test(dp)
shapiro.test(ds)

#QQplot
par(mfrow=c(1,2))
qqnorm(dp,main='placebo')
qqline(dp,col='red')
qqnorm(ds,main='sulindac',ylab=NA)
qqline(ds,col='red')
```
</div>

Calculate the ranks of sample values. Plot the ranks of sample values.

<div class='fold s o'>
```{r,fig.align='center',out.width = '70%'}

ranked_values<-cbind(
  c(rep(1,length(dp)),rep(2,length(ds))),
  rank(c(dp,ds))
)

plot(ranked_values,cex=1.5, xlim=c(0.5, 2.5), axes=FALSE,
     ylab="number of polyps",xlab = NA)

axis(1, c(1, 2), c("patients", "control"))
axis(2)
box()
```
</div>

Perform Mann-Whitney U test, in this case you need to set `correct=FALSE`, because the  continuity correction can not  be applied as there are ties in our sample. Conduct a t-test. Compare results from these two tests.

<div class='fold s o'>
```{r,results='hold',warning=FALSE}
wilcox.test(ds,dp,correct=FALSE)
t.test(ds,dp)
```
</div>
Interpret the results of the test, keeping in mid the analysis made before.



## Exercise 3
Assume there are two different types of disease (type A and B). The mutation rate of a particular gene is $p_1$ under type A and $p_2$ under type B. To compare the mutation rates, tissue samples of $n_A=50$ patients of disease type A and $n_B=100$ patients of disease type B are studied.

Example data is drawn from binomial distributions. Look at the sample values of both patient groups (a value of 0 means no mutation, a value of 1 means there is a mutation).

```{r echo=TRUE}
n1 = 50
n2 = 100
p1 = 0.1
p2 = 0.15
typeA = rbinom(n = n1, prob = p1, size = 1)
typeB = rbinom(n = n2, prob = p2, size = 1)
```

Mutation levels are summarized in a 2x2 cross tabulation:

```{r, echo=TRUE}
mutation = c(typeA, typeB)
group = c(rep("type A", n1), rep("type B", n2))
X = table(mutation, group)
prop.table(X, 2)
```

Perform Fisher's exact test to see whether mutation rates are significantly different between the two groups (**<tt>fisher.test</tt>**).

<div class='fold s o'>
```{r results='hold'}
fisher.test(X)

## Odds ratio
odds_A=p1/(1-p1)
odds_B=p2/(1-p2)
OR=odds_B / odds_A

# Null hypothesis: OR equals to 1
# Null hypothesis : Disease type and mutation status are independent
```
</div>

Change samples sizes $n_1$, $n_2$ and mutations rates $p_1$, $p_2$. Discuss how results change.

<div class='fold s o'>
```{r collapse=TRUE}
n1 = 50
n2 = 100
p1 = 0.1
p2 = 0.5
typeA = rbinom(n = n1, prob = p1, size = 1)
typeB = rbinom(n = n2, prob = p2, size = 1)

mutation = c(typeA, typeB)
group = c(rep("type A", n1), rep("type B", n2))
X = table(mutation, group)

fisher.test(X)

odds_A=p1/(1-p1)
odds_B=p2/(1-p2)
odds_B / odds_A

```
</div>





