---
title: "Hypothesis Tests: Non-normal"
---
# Exercise 5.1 
Assume the expression of a particular gene is measures in blood samples of $n=10$
patients with Parkinson's disease and in $n=10$ healthy controls.

Example data is drawn from exponential distributions as follows:

```{r ex5.1, echo=TRUE, eval=FALSE}
n = 10
patients = rexp(n = n, rate = 1/2)
controls = rexp(n = n, rate = 1/1)
```

Plot of sample values:

```{r ex5.2, echo=TRUE, eval=FALSE}
complete = c(patients, controls)
group = c(rep(1, n), rep(2, n))
plot(group, complete, cex=1.5, xlim=c(0.5, 2.5), axes=FALSE, ylab="gene expression")
axis(1, c(1, 2), c("patients", "control"))
axis(2)
box()
```

Calculate the ranks of sample values. Like above, plot the ranks of sample values.

Use quantile-quantile plots to check for the assumption of normal distribution.

Perform Mann-Whitney U test (Wilcoxon Rank Sum test, R function \texttt{wilcox.test}). Conduct a t-test. Compare results from these two tests.

# Exercise 5.2
Assume there are two different types of disease (type A and B). The mutation rate of a particular gene is \texttt{p1} under type A and \texttt{p2} under type B. To compare the mutation rates, tissue samples of \texttt{n1}=50 patients of disease type A and \texttt{n2}=100 patients of disease type B are studied.

Example data is drawn from binomial distributions. Look at the sample values
of both patient groups - a value of "0" means no mutation, a value of "1" means there is a mutation.

```{r ex5.4, echo=TRUE, eval=FALSE}
n1 = 50
n2 = 100
p1 = 0.1
p2 = 0.15
typeA = rbinom(n = n1, prob = p1, size = 1)
typeB = rbinom(n = n2, prob = p2, size = 1)
```

Mutation levels are summarized in a 2x2 cross tabulation:

```{r ex5.5, echo=TRUE, eval=FALSE}
mutation = c(typeA, typeB)
group = c(rep("type A", n1), rep("type B", n2))
X = table(mutation, group)
X
prop.table(X, 2)
```

Perform Fisher's exact test to see whether mutation rates are significantly
different between the two groups (R function: \texttt{fisher.test}).

Change samples sizes \texttt{n1}, \texttt{n2} and mutations rates \texttt{p1}, \texttt{p2}. Discuss how results change.





