---
title: "Correlation and Regression"
---

# Exercise

## Exercise 6.1 (Correlation)
Assume Proteins A, B and C are members of a cellular pathway that is related to a certain disease. The concentration of these protein was measured in $n=5000$ patients.

Read the corresponding data **<tt>study2.csv</tt>** into the R-environment.

<div class='fold s o'>
```{r}
X = read.table("data/study2.csv", sep=";", header=TRUE)
str(X)
```
</div>

Make scatter plots of each pair of proteins (**<tt>plot</tt>**).

<div class='fold s o'>
```{r fig.align='center',out.width = '70%'}
plot(X)
```
</div>

Determine Pearson's correlation between each pair of proteins (**<tt>cor.test</tt>**).

<div class='fold s o'>
```{r,results='hold'}
cor.test(X$Protein.A, X$Protein.B)
cor.test(X$Protein.A, X$Protein.C)
cor.test(X$Protein.B, X$Protein.C)
```
</div>

## Exercise 6.2 (Linear Regression)

From the correlation analysis, protein A appears to be related to protein B. From earlier research, it is known that protein B depends on protein A. Model the relation between the two proteins using a simple linear regression (**<tt>lm</tt>**).
<div class='fold s'>
```{r}
L1 = lm(X$Protein.B ~ X$Protein.A)
```
</div>

Evaluate the  **<tt>summary</tt>** over the linear model to check the goodness of fit. 

<div class='fold o'>
```{r}
summary(L1)
```
</div>

Do you know the meaning behind each section of summary?

<details><summary>T-statistic</summary>
<p>The t-Test checks the nullity of the coefficient.
</p>
</details>
<details><summary>Multiple R-squared</summary>
<p>Percentage of variability explained by the model.</p>
</details>

<details><summary>F-statistic</summary>
<p>The F-Statistic  checks if at least one of your coefficients are nonzero.
</p>
</details>


The output of the **<tt>lm</tt>** is a list which contains the estimates of the linear regression model. Use the latter to draw 
the regression line into the scatter plot (**<tt>abline</tt>**).

<div class='fold s o'>
```{r}

par(mfrow=c(1,1))
plot(X$Protein.A, X$Protein.B, cex.axis=1.5, cex.lab=1.5, 
     xlab="Concentration of Protein A",
     ylab="Concentration of Protein B")
grid()
#The coefficients of the linear regression model
#L1$coefficients
abline(a=L1$coefficients[1], b=L1$coefficients[2], col=2, lwd=3)
```
</div>