---
title: "Estimation and Confidence Intervals"
---


```{r,echo=FALSE}
op_eval=TRUE
op_echo=TRUE
```

# For loop
A for loop is used to iterate an instruction over a vector. The syntax of a for 
loop in R is:
```{r eval=FALSE}
for (val in sequence)
{
instruction
}
```

where sequence is a vector and val takes on each of its value during the loop. In each iteration the statement is evaluated. Usually the statement depends on val, e.g. `print(val)`. The Fibonacci sequence is defined as :
$$
x_{0}=0 \quad x_{1}=1\\
x_{i}=x_{i-1}+x_{i-2} \quad\forall i>1
$$
Using a for loop, print the first 10 numbers of the Fibonacci sequence.
<div class='fold s o'>
```{r,eval=op_eval,echo=op_echo}
fib<-c(0,1)
# in R the vectors are index starting from 1
for(j in 3:10){
  fib[j]<-fib[j-1]+fib[j-2]
}
print(fib)
```
</div>

# Examples

We want to create a matrix where the first row contains 200 observations taken from exponential distributions with rate $\lambda$ equal to $\frac{1}{16}$, the second with $\frac{1}{22}$, and the third with $\frac{1}{36}$.

<div class='fold s o'>
```{r}
# Vector of rates
r<-c(1/16,1/22,1/36)

# Let's create an empty vector that will store our matrix
M<-c()

# The sequence of this for cycle will be 1,2,3, the indexes of the r vector
for(i in 1:3){
  # At each step we bind a column to M and store the resulting object in M
  M<-cbind(M,rexp(n=100, rate = r[i]))
  # To understand the process we print the first three rows of M at each step
  print(head(M,n=3))
}
```
</div>

For each column, we want to calculate the mean. 

<div class='fold s o'>
```{r}
# The sequence of the for cycle will be 1,2,3 as the matrix M has 3 columns

# We define a vector that contains all the means 
M_col_mean<-c()

for(j in 1:3){
  M_col_mean<-c(M_col_mean,mean(M[,j]))
} 
```
</div>

Knowing that the mean of an exponential distribution of rate $\lambda$ is $\frac{1}{\lambda}$, plot the density function of the three exponential distributions and add to each one the mean and the sample mean.

<div class='fold s o'>
```{r}
par(mfrow=c(2,2)) # Layout with two columns and two rows
x.axis = seq(0, 50, length.out=1000)

for(i in 1:3){
  y.axis = dexp(x.axis, rate=r[i])
  plot(x.axis, y.axis, type="l",
       ylab = '',xlab = '',
       ylim=c(0,0.04))
  # Add the sample mean,red circle
  points(M_col_mean[i], y = 0, pch=16,col='red')
  # Add the mean, blue triangle
  points(1/r[i],y=0,pch=17,col='blue')
}

```
</div>
# Exercises

## Exercise 3.1

Draw 10 random samples of size $N = 5$ from the standard normal distribution ($\mu$ = 0, $\sigma$ = 1),**<tt>rnorm()</tt>**. Calculate the mean of each sample. Calculate the mean of all 10 means.

<div class='fold s o'>
```{r results='hold',eval=op_eval,echo=op_echo}

# Sample creation
M<-rnorm(n = 5, mean=0, sd=1) #variable to store the samples
for(j in 1:9){
  M<-cbind(M,rnorm(n = 5, mean=0, sd=1))
}
# Otherwise you can create an empty matrix 5x10 
# and then use a for loop to fill  it
# M<-matrix(NA, ncol = 10, nrow = 5)
# for(j in 1:10){
#   M[,j]<-rnorm(n = 5, mean=0, sd=1)
# }

# Calculate the mean of each sample

M_col_mean<-c()
for(j in 1:10){
  M_col_mean[j]<-mean(M[,j])
}
# Otherwise using the apply function
# we apply the function mean over each column of the matrix M
#M_col_mean<-apply(M, 2, mean)

# Calculate the mean of the means
M_mean<-mean(M_col_mean)
M_col_mean
M_mean
```
</div>

Plot the density function of the standard normal distribution. Using the function **<tt>points()</tt>** draw the sample means and the mean of means into the plot.

<div class='fold s o'>
```{r fig.align='center',out.width = '80%',eval=op_eval,echo=op_echo}
x.axis = seq(-3, 3, length.out=100)
y.axis = dnorm(x.axis, mean=0, sd=1)
plot(x.axis, y.axis, type="l")
for(i in 1:10) {
  points(M_col_mean[i], y = 0, pch=4)
}
points(M_mean, y = 0, pch=4, col=2, lwd=3)
```
</div>
Repeat the above computer experiment with $N = 25$. 
<div class='fold s o'>
```{r echo=FALSE,fig.align='center',out.width = '80%',eval=op_eval,echo=op_echo}
M<-rnorm(n = 25, mean=0, sd=1) #variable to store the samples
for(j in 1:9){
  M<-cbind(M,rnorm(n = 25, mean=0, sd=1))
}

# Calculate the mean of each sample
M_col_mean<-apply(M, 2, mean)

# Calculate the mean of the means
M_mean<-mean(M_col_mean)

# plot
x.axis = seq(-3, 3, length.out=100)
y.axis = dnorm(x.axis, mean=0, sd=1)
plot(x.axis, y.axis, type="l")
for(i in 1:10) {
  points(M_col_mean[i], y = 0, pch=4)
}
points(M_mean, y = 0, pch=4, col=2, lwd=3)
```
</div>


## Exercise 3.2
A producer gives the information that the content of a pharmaceutical substance in the tablets of some preparation A is normally distributed with an expectation of 500 mg and with a variance  of 10mg. To check this claim, the contents of 25 randomly chosen tablets are determined.

Load the <a href="./data/study3.csv" download><tt>study3.csv</tt></a> data frame and plot the histogram of the data, remember to select a reasonable choice of the parameter **<tt>breaks</tt>**.
<div class='fold s o'>
```{r fig.align='center',out.width = '80%',eval=op_eval,echo=op_echo}
x<-read.table("./data/study3.csv",sep=";",header = TRUE)[[1]] # the values are separated by ; in the csv file
hist(x,breaks = 7)
```
</div>

Determine a 95%-confidence interval for the expectation from the data:

<details><summary>Hint</summary>
<p>
Consider an i.i.d. sample of length  $n$ from $X\sim N(\mu,\sigma)$ then:
$$P\left(\bar{X}-z_{\alpha/2}\frac{\sigma}{\sqrt{n}}\leq \mu\leq \bar{X}+z_{\alpha/2}\frac{\sigma}{\sqrt{n}} \right)=1-\alpha\;,$$
where $z_{\alpha/2}$ is the $\alpha/2$ critical value of a $Z\sim N(0,1)$ distribution, namely $P(Z\geq z_{\alpha})=\alpha$.
</p>
</details>
<p></p>

<div class='fold s o'>
```{r,eval=op_eval,echo=op_echo}
N=25
alpha = 0.05
MEAN = mean(x)
s = sd(x)
SEM = s / sqrt(N) # standard error of the mean
Z = qnorm(1 - alpha / 2, 0, 1)# calculate the critical values
lower_bound<-MEAN - Z * SEM # Lower Bound
upper_bound<-MEAN + Z * SEM # Upper Bound
print(paste0('the confidence interval is [',round(lower_bound,2),',',round(upper_bound,2),']'))
```
</div>


Plot the confidence interval against the histogram, use **<tt>segment</tt>** function.

<div class='fold s o'>
```{r}
hist(x,breaks = 7)
segments(x0=round(lower_bound,2), y0=0, x1 = round(upper_bound,2), y1 = 0,col='red',lwd=5)
```
</div>

Given the confidence intervals, could you trust the producer? Change 

1. the sample size 
2. the mean
3. the standard deviation
4. the level of confidence. 

Discuss changes and results.

Draw 100 random samples of size $n = 25$ from the normal distribution ($\mu$ = 500 mg, $\sigma$ = 10 mg). Calculate
for each one the  95%-confidence interval, how many of them fail to contain the true value?

<div class='fold s o'>
```{r fig.align='center',out.width = '80%',eval=op_eval,echo=op_echo}
# generate the data
M<-rnorm(n = 25, mean=500, sd=10) #variable to store the samples
for(j in 1:99){
  M<-cbind(M,rnorm(n = 25, mean=500, sd=10))
}

M_conf<-matrix(nrow=2,ncol=100)

for(j in 1:100){
  MEAN<-mean(M[,j])
  SEM<-sd(M[,j])/sqrt(25)
  z<-qnorm(1 - 0.05 / 2, 0, 1)
  M_conf[1,j]<-MEAN - Z * SEM # Lower Bound
  M_conf[2,j]<-MEAN + Z * SEM # Upper Bound
}

# percentage of interval that contains the true value for the mean
sum((M_conf[1,]<=500 & 500<=M_conf[2,]))/100

#EXTRA
# usage of if else statement to give the color to the segments
plot.new()
plot(1, type = "n", xlab = "", ylab = "",xlim = c(0, 100), ylim = c(min(M_conf[1,j])-10, max(M_conf[2,j])+10))
for(j in 1:100){
  segments(x0=j,y0=M_conf[1,j],x1=j,y1=M_conf[2,j],col=ifelse((M_conf[1,j]<=500 & 500<=M_conf[2,j]),'black','red'))
}
abline(h=500,col='red')

```
</div>





