---
title: "Analysis of variance"
---
# Theory 

Let's consider the framework in which there are k levels of a factor of interest, which identify k different groups of observations.

The statistical model for one-way analysis of variance is:
$$
y_{ij}=\mu+\tau_{i}+\epsilon_{ij}\quad i=1,\dots,k \text{ and } j=1,\dots,n_{i}
$$
where $\mu$ is the general mean, $\tau_{i}$ is the treatment effect for group $i$ and $\epsilon_{ij}\sim N(0,\sigma^2)$. By considering a suitable F-test is possible to test the null hypothesis that all mean are equal ($\tau_i=0\;\forall i$) against  at least one treatment effect not null. To perform this test we use the **<tt>aov</tt>** function.

# Formulas in R

The the formulas in R are used to specify the symbolic model as well as generating the intended design matrix. The formula are characterized by the $\sim$ as you can see in this example `y~x`.

Let y and x1 be two variables, with `y~x1`we mean that y, the dependent variable, depends on x1, the independent variable. When the dependence relation is quadraticwe will write `y~I(x1^2)`, indeed you nedd to use **<tt>I()</tt>**, as-is operator, to apply a function to a variable. Now let x2 be another variable, with`y~x1+x2`we mean that y depends on x1 and x2 but there is no dependence between x1 and x2. We will use the $-$ sign to ignore an object in the analysis `y~x1+x2`.If we want to express the dependence of y on x1, x2 and their interaction we will write `y~x1+x2+x1:x2`, the former is equal to`y~x1*x2`.

# Exercises

## Exercise 8.1 (One-way analysis of variance)
Assume the concentration of a certain metabolite was measured in three experimental groups.

Sample sizes:
```{r ex1.1}
n1 = 12
n2 = 18
n3 = 8
```

Mean concentration in the population from which samples were drawn:
```{r ex1.2}
mu1 = 1
mu2 = 4.5
mu3 = 6
```

Measurement values can be generated using normal distribution:
```{r ex1.3}
x1 = rnorm(n=n1, mean=mu1, sd=3)
x2 = rnorm(n=n2, mean=mu2, sd=3)
x3 = rnorm(n=n3, mean=mu3, sd=3)
```

a) Plot the concentration versus the group levels (**<tt>boxplot</tt>**).

<div class='fold s o'>
```{r results='hold'}
boxplot(x1, x2, x3, cex.lab=1.5, cex.axis=1.5, xlab="Group", 
        ylab="Concentration (mg/ml)", names=c(1, 2, 3), lwd=2)

x = c(x1, x2, x3)
group = (c(rep('1', n1), rep('2', n2), rep('3', n3)))

aov.oneway = aov(x ~ group)
summary(aov.oneway)

```
</div>

b) Perform one-way analysis of variance (**<tt>aov</tt>**).

What dose the result suggest?

(EXTRA) Plot the stripplot of the the data set to add the mean values of each group and to draw the segments connecting them. To do this, use the **<tt>coefficients</tt>** of the output given by the function **<tt>summary.lm</tt>** applied to the object obtained from **<tt>aov</tt>** .

<div class='fold s o'>
```{r}
plot(x ~ group,
     xlab="Group",
     ylab="Concentration (mg/ml)",
     pch=16)

# mean of the first group
tr1<-summary.lm(aov.oneway)$coefficients[1,1] 
# mean of the second group
tr2<-tr1+summary.lm(aov.oneway)$coefficients[2,1] 
# mean of the second group
tr3<-tr1+summary.lm(aov.oneway)$coefficients[3,1] 

points(c(1,2,3),c(tr1,tr2,tr3),col='red',lwd=2)
lines(c(1,2,3),c(tr1,tr2,tr3),col='red',lwd=2)

```
</div>

c) Perform subsequent pairwise group comparisons by t-tests to see which groups are different. As the chances chances of committing a type I error for a series of comparisons is greater than the error rate for any one comparison alone, use the **<tt>p.adjust</tt>** function  with to **<tt>bonferroni</tt>** method to adjust the p-values.

<div class='fold s o'>
```{r,results='hold'}
(t1 = t.test(x1, x2))
(t2 = t.test(x1, x3))
(t3 = t.test(x2, x3))

## Using bonferroni correction we can adjust the p-values 
p.adjust(p = c(t1$p.value, t2$p.value, t3$p.value), method = "bonferroni")
```
</div>


d) Repeat the exercise with increased variances in the measurement values (e.g. sd = 5).

<div class='fold s o'>
```{r results='hold'}

x1 = rnorm(n=n1, mean=mu1, sd=5)
x2 = rnorm(n=n2, mean=mu2, sd=5)
x3 = rnorm(n=n3, mean=mu3, sd=5)

x = c(x1, x2, x3)
group = (c(rep('1', n1), rep('2', n2), rep('3', n3)))

aov.oneway = aov(x ~ group)
summary(aov.oneway)

```
</div>


## Exercise 8.2 (Repeated measurements analysis of variance)
Some concentration was measured in $n = 10$ patients repeatedly at $T = 4$ time points. Load the dataset **<tt>Repeated_OneWay.csv</tt>**.

```{r}
X <- read.csv("./data/Exercise_8_3_Repeated_OneWay.csv", sep=";")
str(X)
```

a) Plot the concentration across different time-points. Use one-way repeated measures ANOVA to detect whether there is a time effect without the information of repeated measures.

<div class='fold s o'>
```{r}
boxplot(X$Concentration ~ X$Time, 
        xlab="Time", 
        ylab="Concentration")

summary(aov(X$Concentration ~ X$Time))

```
</div>

b) Perform one-way repeated measures ANOVA with information of repeated measures. This is equal to add a random effect to the model that keeps in mind the correlation between observation from the same patient, [random effects model](https://stat.ethz.ch/~meier/teaching/anova/random-and-mixed-effects-models.html#random-effects-models). To do this, wrap the relevant variable inside **<tt>Error()</tt>** operator and add it to the formula.

<div class='fold s o'>
```{r}
summary(aov(X$Concentration ~ X$Time + Error(X$Patient)))
```
</div>
c) A second group was studied in the above study. Plot the concentration versus the group levels across different time-points. Analyse whether there is an effect by time, group or an interaction between time and group (**<tt>boxplot</tt>**). Data is given in the folder
**<tt>Repeated_TwoWay.csv</tt>**.

<div class='fold s o'>
```{r results='hold'}
X = read.csv("./data/Exercise_8_3_Repeated_TwoWay.csv", sep=";")
str(X)

boxplot(X$Concentration ~ X$Time + X$Group, 
        xlab="Time", 
        ylab="Concentration", 
        col=c(2,2,2,2,3,3,3,3)
        )

#3.1 means time 3 group 1

legend("topleft", c("Group 1", "Group 2"), col=c(2,3), ncol=2, pch=15)

```
</div>

d) Conduct Anova without information of repeated measures. Also, conduct Anova with information of repeated measures.

<div class='fold s o'>
```{r results='hold'}
#Without information of repeated measures; and with intreacttions
summary(aov(X$Concentration ~ X$Time + X$Group + X$Time:X$Group))

### With information of repeated measures
summary(aov(X$Concentration ~ X$Time + X$Group + X$Time:X$Group + Error(X$Patient)))
```
</div>
