---
title: "ROC curves"
---

```{r, code=readLines('_setup.R'),echo=FALSE}
```

# Exercises
## Exercise 9.1
The concentration of the s100b-protein was measured in $n=113$ patients. Additionally, the clinical
outcome of patients was classified as either "Good" or "Poor" using a gold standard. Dataset is part of the **<tt>pROC</tt>** package.

a) Install and load the **<tt>pROC</tt>** package. Then load the data set **<tt>aSAH</tt>** from **<tt>pROC</tt>**. This dataset summarizes clinical and laboratory variables of 113 patients with an aneurysmal subarachnoid hemorrhage.

<div class='fold s o'>
```{r warning=FALSE, error=FALSE,message=FALSE,eval=op_eval,echo=op_echo}
#install.packages("pROC")
## Load the library
library(pROC)
## Read the data set
data("aSAH", package = "pROC")
str(aSAH)
```
</div>

b) Perform an ROC analysis using **<tt>roc</tt>** function. 

<div class='fold s o'>
```{r collapse=TRUE,eval=op_eval,echo=op_echo}
roc_s100b = roc(outcome ~ s100b, data = aSAH) #response~predictor
print(roc_s100b)
roc_s100b$sensitivities
```
</div>

The function set automatically the direction of the threshold $t$, namely if `direction: <` the test is positive if s100b$\geq t$ otherwise is positive if s100b$\leq t$.

Using the output of the **<tt>roc</tt>** function, print for each threshold value the sensitivity,the specificity, and the Youden index values, namely create a table that contains the four values and print it.

<div class='fold s o'>
```{r eval=op_eval,echo=op_echo}

result = data.frame(
       'Threshold'= roc_s100b$thresholds,
       'Sensitivity'= roc_s100b$sensitivities,
       'Specificity'= roc_s100b$specificities,
       'Youden'= roc_s100b$sensitivities + roc_s100b$specificities - 1
)


head(result)
```
</div>

Knowing that Specificity=P(Test negative result|Patient healthy),  Sensitivity=P(Test positive| Patient diseased) could you explain the values obtained for $-\infty$ threshold.

<details><summary>Solution</summary>
<p> If the threshold is set to $-\infty$, than all the patients will be labeled as positive, as s100b will be always bigger than $-\infty$.
</p>
</details>
<br>

c) Knowing that the ROC curve is 1-Specificity vs Sensitivity, plot the ROC curve, and interpret the results. Use the **<tt>text</tt>** function to add the AUC value to the plot. Furthermore using the **<tt>points</tt>** function mark up with red color the point corresponding to the maximum Youden index value.

<div class='fold s o'>
```{r results='hold',fig.align='center',out.width = '80%', eval=op_eval,echo=op_echo}
par(mfrow=c(1,2))
stripchart(s100b~outcome,data=aSAH,col='black',vertical=TRUE,pch=19)
plot(x = 1 - roc_s100b$specificities,
     y = roc_s100b$sensitivities, 
     type = "l",
     xlab = "1 - Specificity",
     ylab = "Sensitivity",
     ylim = c(0, 1), 
     xlim = c(0, 1))
lines(x = 0:1, y = 0:1, lty = "dashed")

temp<-length(roc_s100b$specificities)
points(x = 1 - roc_s100b$specificities[seq(2,temp,5)],
       y = roc_s100b$sensitivities[seq(2,temp,5)]) 
text(0.6,0.1,paste('AUC:',round(roc_s100b$auc,2)))

#same result with plot.roc
# plot.roc(roc_s100b)

# Youden index
max_youden<-result[which.max(result$Youden),]
points(x=1-max_youden$Specificity,y=max_youden$Sensitivity,col='red',pch=19)

text(x=1-max_youden$Specificity,y=max_youden$Sensitivity, label=max_youden$Threshold, cex=0.9, font=2,pos=3)
```
</div>


d) Besides s100b, the enzyme NDKA was measured. Make again a ROC analysis and draw the ROC curve on top of the existing plot. You can use **<tt>plot</tt>** and **<tt>lines</tt>** or **<tt>plot.roc</tt>** and **<tt>lines.roc</tt>** functions.

<div class='fold s o'>
```{r warning=FALSE, error=FALSE,message=FALSE,fig.align='center',out.width = '80%', eval=op_eval,echo=op_echo}
plot.roc(outcome ~ ndka, data = aSAH, main="ROC comparison", col= "red")

lines.roc(outcome ~ s100b,data=aSAH, col="blue")
```
</div>


c) Perform a statistical test to see whether s100b or NDKA is a better predictor for the clinical outcome (**<tt>roc</tt>**).

[//]: # (https://statisticaloddsandends.wordpress.com/2020/06/07/what-is-the-delong-test-for-comparing-aucs/)

<div class='fold s o'>
```{r results='hold',warning=FALSE,error=FALSE,message=FALSE,fig.align='center',eval=op_eval,echo=op_echo}

roc_ndka = roc(outcome ~ ndka, data = aSAH)
roc_s100b$auc
roc_ndka$auc

roc.test(roc_s100b, roc_ndka)
```
</div>
