---
title: "Data management and descriptive methods"
---

```{r, code=readLines('_setup.R')}
```

CASE STUDY: In a clinical trial 12 patients are randomly assigned to two different treatments. Blood measurements are taken before and after the treatment.
Load the **<tt>study1.csv</tt>** data frame into the R workspace.

```{r}
study1<-read.table("./data/study1.csv",sep=";",header = TRUE) # the values are separated by ; in the csv file
```



# Exercises

## Execrise 1.1
Calculate mean, median, variance, standard deviation, quartiles and the sum of the measurements before treatment (**<tt>mean()</tt>**).

<div class='fold s o'>
```{r collapse=TRUE,eval=op_eval,echo=op_echo}
X<-study1$before
# we will show the direct function and the calculation behind it
# sum of the elements of the vector
sum(X)
# the sample mean
sum(X)/length(X)
mean(X)
# the sample variance
sum((X-mean(X))^2)/(length(X)-1)
var(X)

# the sample standard deviation
sqrt(var(X))
sd(X)

#the median
mean(sort(X)[c(length(X)/2,length(X)/2+1)])
median(X)

# quantile
quantile(X, probs=c(0.25, 0.50, 0.75))
# A quartile is the value of the quantile at the probabilities 0.25, 0.5 and 0.75.

# all of this can be achieved easily using the summary function
summary(X)
```
</div>


## Exercise 1.2

Plot a boxplot and a histogram of the measurements obtained before the treatment. 
Correct the title of the axes. Choose appropriate range of axes (**<tt>boxplot()</tt>**).

<div class='fold s o'>
```{r, fig.align='center',eval=op_eval,echo=op_echo}
par(mfrow=c(1,2))# define the plot area one row, two columns
X<-study1$before
boxplot(X, 
        xlab = "Before treatment", 
        ylab = "Blood measurements", 
        ylim = c(0.2, 0.7), 
        main = "Box plot", 
        horizontal = TRUE)
points(X, y = rep(0.6, 12), pch=1,
       col="orange", cex=0.75)
# Histogram
hist(X, 
     xlim = c(0.2, 0.7), 
     xlab = "Before treatment", 
     main = "Histogram", 
     breaks = 12)
points(X, rep(0, 12), pch=15, col="orange", cex=0.75)
```
</div>



## Exercise 1.3

Determine the absolute and relative frequencies of patients in the study groups(**<tt>table()</tt>**). Plot  a bar chart and a pie chart of the parameter "group" (**<tt>pie()</tt>** and **<tt>barplot()</tt>**).

<div class='fold s o'>
```{r, fig.align='center',results='hold',eval=op_eval,echo=op_echo}
X<-study1$group
## absolute and relative 
table(X); table(X) / length(X)
```
</div>
<div class='fold s o'>
```{r, fig.align='center',results='hold',eval=op_eval,echo=op_echo}
par(mfrow=c(1,2))# define the plot area one row, two columns
pie(table(X),
    main = "Pie chart"
    )
barplot(table(X),
        main = "Bar chart"
        )
```
</div>

