---
title: "Analysis of variance"
---

# Exercise 8.1 (One-way analysis of variance)
Assume the concentration of a certain metabolite was measured in three experimental groups.

Sample sizes:
```{r ex1.1, echo=TRUE, eval=FALSE}
n1 = 12
n2 = 18
n3 = 8
```

Mean concentration in the population from which samples were drawn:
```{r ex1.2, echo=TRUE, eval=FALSE}
mu1 = 4
mu2 = 4.5
mu3 = 6
```

Measurement values can be generated using normal distribution:
```{r ex1.3, echo=TRUE, eval=FALSE}
x1 = rnorm(n=n1, mean=mu1, sd=3)
x2 = rnorm(n=n2, mean=mu2, sd=3)
x3 = rnorm(n=n3, mean=mu3, sd=3)
```

a) Plot the concentration versus the group levels and perform one-way analysis of variance (\texttt{boxplot}, \texttt{aov}).

b) Perform subsequent pairwise group comparisons by t-tests to see which groups are different.

c) Repeat the exercise with increased variances in the measurement values (e.g. sd = 5).


# Exercise 8.2 (Repeated measurements analysis of variance)
Some concentration was measured in $n = 10$ patients repeatedly at $T = 4$ time points. Load the dataset \texttt{Exercise\_8\_2\_Repeated\_OneWay.csv}.

a) Plot the concentration across different time-points. Use one-way repeated measures ANOVA to detect whether there is a time effect without the information of repeated measures (\texttt{aov}).

b) Perform one-way repeated measures ANOVA with information of repeated measures.

c) A second group was studied in the above study. Plot the concentration versus the group levels across different time-points. Analyse whether there is an effect by time, group or an interaction between time and group. Data is given in the folder \texttt{Exercise\_8\_3\_Repeated\_TwoWay.csv}.

d) Conduct Anova without information of repeated measures. Also, conduct Anova with information of repeated measures.

